---
title: "Identification of dominant TCRs following co-culture with TMG-loaded APCs"
author: "Endika Prieto <endikaprieto@vhio.net>"
date: "2026-04-24"
---

# setup

```{r, warning=FALSE, message=FALSE}
# --- Load libraries and custom functions ---
set.seed(123)
source("/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/pbl/single_cell/sc_pipeline/scripts/TIL15/setup/setup-m1.R")

# --- Define paths and sample identifiers ---
datapath <- "/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/pbl/single_cell/sc_pipeline/patients"
sample <- "TIL15"
folder_data <- "ITAG_AGG_S02"

# --- Define data loading path ---
datapath_load <- "/Volumes/datos_lab/single_cell"
datapath_mixcr <- "/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/pbl/single_cell/mixcr"

# --- Initialize lists for plot storage ---
plots_list <- list()
pdfcluster_plots <- list()
```

# data load

Assessment of ambient RNA contamination using SoupX.

https://cran.r-project.org/web/packages/SoupX/vignettes/pbmcTutorial.html
https://www.sc-best-practices.org/preprocessing_visualization/quality_control.html
https://cellgeni.github.io/notebooks/html/new-10kPBMC-SoupX.html

```{r}
object_files <- list.dirs(
  file.path(datapath_load, sample, folder_data),
  full.names = TRUE,
  recursive = FALSE
)
object_tags <- c("Sample1", "Sample2", "Sample3", "Sample4")
obj_VH <- list()
soupX_VH <- list()
for (i in seq_along(object_files)) {
  toc <- Read10X(file.path(object_files[i], "sample_filtered_feature_bc_matrix"))
  toc_seurat <- CreateSeuratObject(toc)
  obj_VH[[as.character(object_tags[i])]] <- toc_seurat
  obj_VH[[as.character(object_tags[i])]]$tag <- paste0(object_tags[i], "_")
  tod <- Read10X(file.path(object_files[i], "sample_raw_feature_bc_matrix"))
  sc <- SoupChannel(tod, toc)
  toc_seurat <- PercentageFeatureSet(toc_seurat, pattern = "^MT-", col.name = "percent.mt")
  toc_seurat <- SCTransform(toc_seurat, vars.to.regress = "percent.mt", verbose = FALSE)
  toc_seurat <- RunPCA(toc_seurat, verbose = FALSE)
  toc_seurat <- RunTSNE(toc_seurat, dims = 1:10, seed.use = 123)
  toc_seurat <- RunUMAP(toc_seurat, dims = 1:10, seed.use = 123)
  toc_seurat <- FindNeighbors(toc_seurat, dims = 1:10, verbose = FALSE)
  toc_seurat <- FindClusters(toc_seurat, verbose = FALSE, resolution = 1.2, random.seed = 123)
  DefaultAssay(toc_seurat) <- "RNA"
  meta <- toc_seurat@meta.data
  umap <- toc_seurat@reductions$umap@cell.embeddings
  sc <- setClusters(sc, setNames(meta$seurat_clusters, rownames(meta)))
  sc <- setDR(sc, umap)
  sc  <- autoEstCont(sc, forceAccept = TRUE)
  soupX_VH[[as.character(object_tags[i])]] <- sc
  rm(toc, toc_seurat, tod, sc, meta, umap, i)
}
dir.create(file.path(datapath, sample, folder_data, "seurat"), recursive = F)
saveRDS(soupX_VH, file.path(datapath, sample, folder_data, "seurat/soupX.rds"))
rm(object_files, object_tags)
```

Gene-level contributions to ambient RNA estimated by SoupX (no correction applied).

```{r}
plots_soupX <- list()
for (i in names(soupX_VH)) {
  data_i <- head(
    soupX_VH[[i]]$soupProfile[
      order(soupX_VH[[i]]$soupProfile$est, decreasing = TRUE), 
    ],
    n = 50
  )
  data_i$gene <- as.character(row.names(data_i))
  data_i$gene <- factor(data_i$gene, levels = data_i$gene)
  data_i$source <- ifelse(
    grepl("^MT-", data_i$gene), "mito",
    ifelse(grepl("^RPS|^RPL", data_i$gene), "ribo", "other")
  )
  plot_i <- ggplot(data_i, aes(x = gene, y = est, fill = source)) +
    geom_col(width = 0.8) +
    geom_text(
      aes(label = ifelse(source == "other", as.character(gene), "")),
      hjust = -0.1, vjust = -0.5,
      size = 2.5,
      angle = 45
    ) +
    coord_cartesian(ylim = c(0, 0.13), clip = "off") +
    scale_fill_manual(
      values = c(
        "mito" = "red3",
        "ribo" = "gray60",
        "other" = "gray20"
      )
    ) +
    labs(
      title = i,
      x = NULL,
      y = "SoupX estimate",
      fill = "Source"
    ) +
    theme_ic +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank()
    )
  plots_soupX[[as.character(i)]] <- plot_i
}
plots_list[["IC01_soupX"]] <- wrap_plots(plots_soupX, ncol = 2)
print(plots_list$IC01_soupX)
rm(soupX_VH, plots_soupX, i, data_i, plot_i)
```

Count distribution across samples.
Broadly consistent, with minor differences in spread and tail length.

```{r}
names <- names(obj_VH)
gene <- "CD3E"
gpha <- lapply(
  setNames(names, names),
  function(x) {
    df <- data.frame(counts = obj_VH[[x]]@assays$RNA$counts[gene, ])
    ggplot(df, aes(x = counts)) +
      geom_histogram(
        bins = 50,
        fill = "gray50",
        color = "black",
        linewidth = 0.2
      ) +
      ggtitle(x) +
      scale_x_continuous(
        breaks = seq(0, max(df$counts, na.rm = TRUE), by = 10),
        expand = expansion(mult = c(0, 0.02))
      ) +
      labs(
        x = "Counts",
        y = "Cells"
      ) +
      theme_ic
  }
)
plots_list[["IC02_counts"]] <- wrap_plots(gpha, ncol = 2)
print(plots_list$IC02_counts)
rm(gpha, gene)
```

VDJ data loading with djvdj.

https://rnabioco.github.io/djvdj/

```{r}
vdj_files <- list.files(file.path(datapath_load, sample, folder_data))[-1]
vdj_dirs <- paste0(file.path(datapath_load, sample, folder_data), "/", vdj_files)
vdj_dirs <- setNames(vdj_dirs, vdj_files)
for (i in seq_along(vdj_dirs)){
  so <- obj_VH[[i]]
  so <- import_vdj(so, vdj_dir = vdj_dirs[[i]])
  obj_VH[[i]] <- so
}
rm(vdj_files, vdj_dirs, so, datapath_load, i)
```

Sample annotation by time point.

```{r}
for (nm in names(obj_VH)) {
  obj <- obj_VH[[nm]]
  obj$sample <- nm
  new_names <- paste0(obj$tag, colnames(obj))
  obj <- RenameCells(obj, new.names = new_names)
  obj_VH[[nm]] <- obj
}
rm(obj, nm, new_names)
```

Sample merging.

```{r}
obj <- c(obj_VH)
obj <- Reduce(merge, obj)
obj <- JoinLayers(obj)
obj$tag <- gsub("_$", "", obj$tag)
obj@meta.data$tag <- factor(obj@meta.data$tag, levels = c("Sample1", "Sample2", "Sample3", "Sample4"))
rm(obj_VH)
```

Quality control overview.

```{r, warning=FALSE}
# --- Logreads ---
obj$logreads <- log(colSums(obj$RNA))
# --- Mito and ribo genes ---
obj <- PercentageFeatureSet(obj, pattern = "^MT-", col.name = "percent.mt")
obj <- PercentageFeatureSet(obj, pattern="^RPL", col.name = "percent.rpl")
obj <- PercentageFeatureSet(obj, pattern="^RPS", col.name = "percent.rps")
obj <- PercentageFeatureSet(obj, pattern="^RPS|^RPL", col.name = "percent.rpa")
# --- TCR genes---
obj <- PercentageFeatureSet(obj, pattern="^TRAV", col.name = "TRAV.avg")
obj <- PercentageFeatureSet(obj, pattern="^TRAJ", col.name = "TRAJ.avg")
obj <- PercentageFeatureSet(obj, pattern="^TRBV", col.name = "TRBV.avg")
obj <- PercentageFeatureSet(obj, pattern="^TRBJ", col.name = "TRBJ.avg")
obj <- PercentageFeatureSet(obj, pattern="^TRGV", col.name = "TRGV.avg")
obj <- PercentageFeatureSet(obj, pattern="^TRDV", col.name = "TRDV.avg")
```

Definition of TCR-positive and TCR-negative cell subsets.

```{r}
md <- obj@meta.data
n1 <- rownames(subset(md, subset = !is.na(cdr3)))
n2 <- rownames(subset(md, subset = is.na(cdr3)))
obj$tag.tcr <- ifelse(colnames(obj) %in% n1, "yes", "no")
```

Comparison of quality control metrics by TCR status.

```{r}
metrics <- c("percent.mt", "percent.rpa", "TRAV.avg", "TRBV.avg")
# --- Cells with TCR ---
qc_tcr <- list()
for (i in metrics) {
  plot_i <- ggplot(
    md[n1, , drop = FALSE],
    aes(x = tag, y = .data[[i]], fill = tag)
  ) +
    geom_boxplot(
      outlier.alpha = 0.2,
      linewidth = 0.3
    ) +
    scale_fill_brewer(palette = "Paired") +
    labs(
      y = i
    ) +
    theme_ic +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )
  if (i == "percent.mt") {
    plot_i <- plot_i +
      geom_hline(
        yintercept = 10,
        linetype = "dashed",
        color = "black",
        linewidth = 0.3
      )
  }
  qc_tcr[[i]] <- plot_i
}
# --- Cells without TCR ---
qc_no_tcr <- list()
for (i in metrics) {
  plot_i <- ggplot(
    md[n2, , drop = FALSE],
    aes(x = tag, y = .data[[i]], fill = tag)
  ) +
    geom_boxplot(
      outlier.alpha = 0.2,
      linewidth = 0.3
    ) +
    scale_fill_brewer(palette = "Paired") +
    labs(
      y = i
    ) +
    theme_ic +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )
  if (i == "percent.mt") {
    plot_i <- plot_i +
      geom_hline(
        yintercept = 10,
        linetype = "dashed",
        color = "black",
        linewidth = 0.3
      )
  }
  qc_no_tcr[[i]] <- plot_i
}
# --- Plots ---
plots_list[["IC03_qc_tcr"]] <- wrap_plots(qc_tcr, ncol = 2) +
  plot_annotation(title = "Cells with TCR")
plots_list[["IC03_qc_no_tcr"]] <- wrap_plots(qc_no_tcr, ncol = 2) +
  plot_annotation(title = "Cells without TCR")
print(plots_list[["IC03_qc_tcr"]])
print(plots_list[["IC03_qc_no_tcr"]])
rm(i, metrics, plot_i, qc_tcr, qc_no_tcr)
```

Evaluation of mitochondrial and ribosomal filtering thresholds.

```{r}
metrics <- c("percent.mt", "percent.rpl", "percent.rps", "percent.rpa")
# --- Cells with TCR ---
qc_curves_tcr <- list()
for (i in metrics) {
  thresholds <- seq(0, 70, by = 1)
  results <- data.frame(
    threshold = numeric(),
    passing_cells = numeric(),
    tag = character()
  )
  for (j in unique(md$tag)) {
    tag_data <- md[n1, , drop = FALSE]
    tag_data <- tag_data[tag_data$tag == j, , drop = FALSE]
    if (nrow(tag_data) == 0) next
    for (k in thresholds) {
      percent_passing <- mean(tag_data[[i]] <= k, na.rm = TRUE)
      results <- rbind(
        results,
        data.frame(
          threshold = k,
          passing_cells = percent_passing,
          tag = j
        )
      )
    }
  }
  results$tag <- factor(results$tag, levels = unique(md$tag))
  plot_i <- ggplot(results, aes(x = threshold, y = passing_cells, color = tag)) +
    geom_line(linewidth = 0.6) +
    scale_color_brewer(palette = "Paired") +
    scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.02))) +
    labs(
      x = paste0("max_", i),
      y = "Fraction passing filter",
      color = "Sample"
    ) +
    theme_ic
  if (i == "percent.mt") {
    plot_i <- plot_i +
      geom_vline(
        xintercept = 10,
        linetype = "dashed",
        linewidth = 0.3,
        color = "black"
      )
  }
  qc_curves_tcr[[i]] <- plot_i
}
# --- Cells without TCR ---
qc_curves_no_tcr <- list()
for (i in metrics) {
  thresholds <- seq(0, 70, by = 1)
  results <- data.frame(
    threshold = numeric(),
    passing_cells = numeric(),
    tag = character()
  )
  for (j in unique(md$tag)) {
    tag_data <- md[n2, , drop = FALSE]
    tag_data <- tag_data[tag_data$tag == j, , drop = FALSE]
    if (nrow(tag_data) == 0) next
    for (k in thresholds) {
      percent_passing <- mean(tag_data[[i]] <= k, na.rm = TRUE)
      results <- rbind(
        results,
        data.frame(
          threshold = k,
          passing_cells = percent_passing,
          tag = j
        )
      )
    }
  }
  results$tag <- factor(results$tag, levels = unique(md$tag))
  plot_i <- ggplot(results, aes(x = threshold, y = passing_cells, color = tag)) +
    geom_line(linewidth = 0.6) +
    scale_color_brewer(palette = "Paired") +
    scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.02))) +
    labs(
      x = paste0("max_", i),
      y = "Fraction passing filter",
      color = "Sample"
    ) +
    theme_ic
  if (i == "percent.mt") {
    plot_i <- plot_i +
      geom_vline(
        xintercept = 10,
        linetype = "dashed",
        linewidth = 0.3,
        color = "black"
      )
  }
  qc_curves_no_tcr[[i]] <- plot_i
}
# --- Plots ---
plots_list[["IC04_qc_mt_filt_tcr"]] <- wrap_plots(qc_curves_tcr, ncol = 2) +
  plot_annotation(title = "Cells with TCR")
plots_list[["IC04_qc_mt_filt_no_tcr"]] <- wrap_plots(qc_curves_no_tcr, ncol = 2) +
  plot_annotation(title = "Cells without TCR")
print(plots_list[["IC04_qc_mt_filt_tcr"]])
print(plots_list[["IC04_qc_mt_filt_no_tcr"]])
rm(
  md, n1, n2, i, j, k, metrics, thresholds, results, tag_data, percent_passing,
  plot_i, qc_curves_tcr, qc_curves_no_tcr
)
```

Stratification of cells by mitochondrial content.

```{r}
obj$tag.mt <- case_when(
  obj$percent.mt > 20 ~ "elim_mt",
  obj$percent.mt >= 15 & obj$percent.mt <= 20 ~ "high_mt",
  obj$percent.mt >= 10 & obj$percent.mt < 15 ~ "int_mt",
  obj$percent.mt < 10 ~ "low_mt"
)
obj$tag.mt <- factor(obj$tag.mt, levels = c("elim_mt", "low_mt", "int_mt", "high_mt"))
# --- Summary table ---
print(table(obj$tag.mt, obj$tag))
```

Association between mitochondrial/ribosomal content and TCR detection.

```{r}
md <- obj@meta.data
md$tag <- factor(md$tag, levels = c("Sample1", "Sample2", "Sample3", "Sample4"))
md$tag.tcr_label <- ifelse(md$tag.tcr == "yes", "Cells with TCR", "Cells without TCR")
md$tag.tcr_label <- factor(
  md$tag.tcr_label,
  levels = c("Cells with TCR", "Cells without TCR")
)
mito_plots_filter <- ggplot(
  md,
  aes(
    x = log(percent.mt),
    y = log(percent.rpa),
    color = tag.mt,
    shape = tag.tcr
  )
) +
  geom_point(alpha = 0.3, size = 1.5) +
  facet_grid(
    rows = vars(tag.tcr_label),
    cols = vars(tag)
  ) +
  geom_vline(
    xintercept = log(c(10, 15, 20)),
    linetype = "dashed",
    linewidth = 0.3,
    color = "black"
  ) +
  scale_color_manual(
    values = c(
      "elim_mt" = "red3",
      "high_mt" = "orange2",
      "int_mt" = "green4",
      "low_mt" = "gray20"
    )
  ) +
  scale_shape_manual(values = c("yes" = 1, "no" = 3)) +
  labs(
    title = "Mitochondrial filtering",
    x = "log(percent.mt)",
    y = "log(percent.rpa)",
    color = "MT category",
    shape = "TCR"
  ) +
  theme_ic
plots_list[["IC05_qc_mito_filter_tcrb"]] <- mito_plots_filter
print(plots_list[["IC05_qc_mito_filter_tcrb"]])
rm(mito_plots_filter, md)
```

Doublet detection using scDblFinder.

https://github.com/plger/scDblFinder?tab=readme-ov-file
https://plger.github.io/scDblFinder/articles/scDblFinder.html

```{r, warning=FALSE}
set.seed(123)
# --- Transform Seurat to SCE ---
sce_obj <- as.SingleCellExperiment(obj, assay = "RNA")
colData(sce_obj)$tag <- obj$tag
sce_obj <- scDblFinder(
  sce_obj,
  samples = "tag",
  clusters = FALSE,
  BPPARAM = BiocParallel::SerialParam(RNGseed = 123)
)
# --- Return to Seurat ---
obj$scDblFinder.score <- sce_obj$scDblFinder.score
obj$scDblFinder.class <- sce_obj$scDblFinder.class
table(obj$scDblFinder.class)
rm(sce_obj)
```

Association between doublet scores and logreads.

```{r}
doublet_plots <- ggplot(
  obj@meta.data,
  aes(x = logreads, y = scDblFinder.score, color = scDblFinder.class)
) +
  geom_point(alpha = 0.3, size = 1.5) +
  facet_wrap(~tag) +
  scale_color_manual(
    values = c(
      "doublet" = "red3",
      "singlet" = "gray20"
    )
  ) +
  labs(
    title = "Doublets",
    x = "logreads",
    y = "scDblFinder score",
    color = "scDblFinder class"
  ) +
  theme_ic
plots_list[["IC06_qc_doublets_tcrb"]] <- doublet_plots
print(plots_list$IC06_qc_doublets_tcrb)
rm(doublet_plots)
```

# mixcr

Cell Ranger did not recover TCR information for a subset of T cells with high TCR gene expression. To rescue these assignments, we used MiXCR. Based on discussions with the MiXCR bioinformatics team, we adopted the following strategy:

1. Assign each cell to its corresponding GEM using GEX-based cell calls generated by Cell Ranger.  
2. Restrict MiXCR analysis to GEMs identified as containing cells.  

This workflow requires the generation of a `10x-sample-sheet.tsv` file, which must be available in the working directory prior to execution.

```{r}
table <- data.frame(
  Sample     = obj@meta.data$tag,
  tagPattern = "",
  `CELL*`    = sub(
    "-1$", "",
    sub(".*_", "", row.names(obj@meta.data))
  ),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
write.table(table, file.path(datapath_mixcr, sample, folder_data, "10x-sample-sheet.tsv"), sep = "\t", row.names = FALSE, quote = FALSE)
rm(table)
```

MiXCR was executed on the selected GEMs.

Resulting clonotypes were loaded into R for downstream analysis.

```{r, warning=FALSE}
files <- c(
  "sample_result.Sample1.clones.tsv",
  "sample_result.Sample2.clones.tsv",
  "sample_result.Sample3.clones.tsv",
  "sample_result.Sample4.clones.tsv"
)
mixcr_list <- list()
for (i in files) {
mixcr <- read.table(file.path(datapath_mixcr, sample, folder_data, i), header = TRUE, sep = "\t", stringsAsFactors = FALSE) %>%
  filter(cellGroup != "contamination") %>%
  select(tagValueCELL, bestVHit, bestDHit, bestJHit, aaSeqCDR3, topChains) %>%
  mutate(
    bestVHit = sub("\\*.*", "", bestVHit),
    bestDHit = sub("\\*.*", "", bestDHit),
    bestJHit = sub("\\*.*", "", bestJHit)
  )
mixcr <- mixcr %>%
  group_by(tagValueCELL) %>%
  summarise(
    ord = order(factor(topChains, levels = c("TRA","TRB"))),
    bestVHit = paste(bestVHit[ord], collapse = ";"),
    bestJHit = paste(bestJHit[ord], collapse = ";"),
    aaSeqCDR3 = paste(aaSeqCDR3[ord], collapse = ";"),
    topChains = paste(topChains[ord], collapse = ";"),
    .groups = "drop"
  ) %>%
  select(-ord) %>%
  distinct()
colnames(mixcr) <- c("barcode", paste0(c("v_gene", "j_gene", "cdr3", "chains"), "_mixcr"))
mixcr <- mixcr %>% filter(if_any(c(v_gene_mixcr, j_gene_mixcr, cdr3_mixcr, chains_mixcr), ~ . != "" & !is.na(.)))
mixcr_list[[i]] <- mixcr
}
mixcr <- bind_rows(mixcr_list) %>% as.data.frame()
# Internal control
barcodes <- read.table(file.path(datapath_mixcr, sample, folder_data, "10x-sample-sheet.tsv"), header = TRUE, sep = "\t", stringsAsFactors = FALSE)
barcodes <- barcodes$CELL.
message(
  round(sum(mixcr$barcode %in% barcodes) / length(mixcr$barcode) * 100, 2),
  "% of mixcr barcodes are in the Seurat object"
)
rm(mixcr_list, barcodes, files, i)
```

MiXCR-derived clonotypes were added to metadata.
 
```{r}
obj@meta.data$barcode <- str_extract(rownames(obj@meta.data), "(?<=_)[^-]+")
row.names <- rownames(obj@meta.data)
obj@meta.data <- obj@meta.data %>% left_join(mixcr, by = "barcode")
obj@meta.data$tag.tcr_mixcr <- ifelse(is.na(obj@meta.data$chains_mixcr) | obj@meta.data$chains_mixcr == "", "no", "yes")
rownames(obj@meta.data) <- row.names
rm(mixcr, row.names)
```

A combined clonotype category (Cell Ranger + MiXCR) was subsequently defined to maximize TCR recovery.

```{r}
obj@meta.data <- obj@meta.data  %>%
  mutate(
    v_gene_combi = ifelse(tag.tcr == "yes", v_gene,
                          ifelse(tag.tcr_mixcr == "yes", v_gene_mixcr, NA_character_)),
    j_gene_combi = ifelse(tag.tcr == "yes", j_gene,
                          ifelse(tag.tcr_mixcr == "yes", j_gene_mixcr, NA_character_)),
    cdr3_combi   = ifelse(tag.tcr == "yes", cdr3,
                          ifelse(tag.tcr_mixcr == "yes", cdr3_mixcr, NA_character_)),
    chains_combi = ifelse(tag.tcr == "yes", chains,
                          ifelse(tag.tcr_mixcr == "yes", chains_mixcr, NA_character_))
  )
obj@meta.data$tag.tcr_combi <- ifelse(is.na(obj@meta.data$chains_combi) | obj@meta.data$chains_combi == "", "no", "yes")
```

Cell filtering, based on:

- Doublets.
- Ribosomal genes.
- TCR genes (https://github.com/ncborcherding/Trex/).
- Genes with very low expression.
- Cells with >20% percent.mt.
- Cells without TCR information.

```{r}
obj <- subset(obj, subset = scDblFinder.class != "doublet")
all_genes <- rownames(obj)
counts <- GetAssayData(obj, assay = "RNA", layer = "counts")
tcr_genes <- setdiff(all_genes, quietTCRgenes(all_genes))
ribo_genes <- quietRibogenes(all_genes)
expressed_genes <- rownames(counts)[rowSums(counts > 0) >= 5]
keep_genes <- setdiff(intersect(all_genes, expressed_genes), c(tcr_genes, ribo_genes))
keep_cells <- rownames(obj@meta.data)[
  obj@meta.data$tag.mt != "elim_mt" &
  obj@meta.data$tag.tcr_combi != "no"
]
obj <- subset(obj, features = keep_genes, cells = keep_cells)
rm(all_genes, counts, tcr_genes, ribo_genes, expressed_genes, keep_genes, keep_cells)
```

# SCTransform

Normalization and variance stabilization were performed using SCTransform as implemented in Seurat.

https://satijalab.org/seurat/articles/sctransform_vignette

```{r}
obj <- SCTransform(obj = obj, vars.to.regress = "percent.mt", verbose = FALSE, return.only.var.genes = FALSE)
obj <- RunPCA(obj, verbose = FALSE)
ElbowPlot(obj, reduction = "pca")
ndim <- 6
obj <- RunTSNE(obj, dims = 1:ndim, seed.use = 123)
obj <- RunUMAP(obj, dims = 1:ndim, seed.use = 123)
obj <- FindNeighbors(obj, dims = 1:ndim, verbose = FALSE)
obj <- FindClusters(obj, verbose = FALSE, resolution = 0.75, random.seed = 123)
rm(ndim)
```

Internal control.

```{r}
g1 <- DimPlot(obj, group.by = "tag.tcr", pt.size = 0.1, cols = c("yes" = "#3c4b99",  "no" = "lightgray")) + theme(aspect.ratio = 1) + ggtitle("cellranger")
g2 <- DimPlot(obj, group.by = "tag.tcr_mixcr", pt.size = 0.1, cols = c("yes" = "#3c4b99",  "no" = "lightgray")) + theme(aspect.ratio = 1) + ggtitle("mixcr")
g3 <- DimPlot(obj, group.by = "tag.tcr_combi", pt.size = 0.15, cols = c("yes" = "#3c4b99",  "no" = "lightgray")) + theme(aspect.ratio = 1) + ggtitle("cellranger+mixcr")
print(wrap_plots(g1, g2, g3, ncol = 3))
rm(g1, g2, g3)
```

# magic

To reduce sparsity and improve signal recovery, data imputation was performed using MAGIC. The Python package `magic-impute` was installed in a dedicated virtual environment and interfaced through `reticulate`.

```{bash}
#cd /Users/endikaprieto/venv
#pip3.12 install magic-impute

#library(reticulate)
#reticulate::use_virtualenv("/Users/endikaprieto/venv", required = TRUE)
#reticulate::py_install(c('magic'))
#reticulate::import("magic")
```

MAGIC was applied to the dataset to obtain imputed expression values.

```{r, warning=FALSE}
obj@assays$MAGIC_SCT <- magic(obj, verbose=F, assay = "SCT", seed = 123)@assays$MAGIC_SCT
```

MAGIC-imputed expression values were incorporated into the Seurat object metadata, and the corresponding MAGIC object was retained for subsequent analyses.

```{r}
# --- Add Magic data to meta.data ---
genes <- c("CD4", "CD8A", "CD8B", "ENTPD1", "PDCD1")
common_cells <- intersect(colnames(obj), colnames(obj@assays$MAGIC_SCT$data))
genes_present <- genes[genes %in% rownames(obj@assays$MAGIC_SCT$data)]
genes_missing <- setdiff(genes, genes_present)
for (g in genes_present) {
  vals <- obj@assays$MAGIC_SCT$data[g, common_cells]
  new_col <- paste0(g, "_mag")
  obj[[new_col]] <- NA_real_
  obj@meta.data[common_cells, new_col] <- as.numeric(vals)
}
# --- Umap plots ---
p <- FeaturePlot(
  obj,
  features = c("PDCD1_mag", "ENTPD1_mag", paste0(genes, "_mag")),
  ncol = 6, pt.size = 0.25
) &
  scale_color_gradientn(
    colours = c("#D9F2E6", "#52B788", "#2C7FB8", "#08306B"),
    oob = scales::squish,
    na.value = "grey80"
  ) &
  theme(aspect.ratio = 1)
plots_list[["IC07_GOIs_Magic"]] <- p
# --- Save Magic data ---
obj_MAGIC <- CreateSeuratObject(obj@assays$MAGIC_SCT)
obj_MAGIC <- AddMetaData(obj = obj_MAGIC, metadata = obj@meta.data)
saveRDS(obj_MAGIC, file.path(datapath, sample, folder_data, "seurat/magic.rds"))
rm(obj_MAGIC)
# --- Eliminate Magic assay and free unused memory ---
DefaultAssay(obj) <- "SCT"
obj[["MAGIC_SCT"]] <- NULL
gc()
rm(g, p, new_col, genes, common_cells, genes_present, genes_missing)
```

Ridge plots illustrating CD8A, CD8B, CD4, ENTPD1, and PDCD1 expression across samples were generated. The distribution valley separating positive and negative expression states was consistent across samples, including both bulk and PD-1–enriched conditions.

```{r, warning=FALSE}
markers <- c("CD8A_mag", "CD8B_mag", "CD4_mag", "ENTPD1_mag", "PDCD1_mag")
ridge_plots <- list()
for (i in markers) {
  x <- data.frame(
    marker_value = obj@meta.data[[i]],
    sample_type = obj@meta.data$tag
  )
  g <- ggplot(x, aes(x = marker_value, y = sample_type, fill = sample_type)) +
    geom_density_ridges(
      aes(point_color = 1, point_fill = sample_type),
      alpha = 0.75,
      jittered_points = TRUE,
      point_alpha = 0.1,
      point_color = "black",
      position = position_points_jitter(width = 0.05, height = 0),
      point_shape = "|"
    ) +
    scale_fill_brewer(palette = "Paired") +
    labs(
      title = i,
      x = "MAGIC-imputed expression",
      y = "Sample",
      fill = "Sample"
    ) +
    theme_ic +
    theme(
      legend.position = "none"
    )
  ridge_plots[[i]] <- g
}
plots_list[["IC08_cd48_density"]] <- wrap_plots(ridge_plots, ncol = 3)
print(plots_list[["IC08_cd48_density"]])
rm(markers, ridge_plots, i, x, g)
```

# pdfCluster

Thresholds for CD4 and CD8A/CD8B expression were defined using density-based clustering (pdfCluster) applied to MAGIC-imputed values.

Deterministic, to avoid randomnes and ensure reproducibility.

```{r}
n_per_sample <- 750
sample_var <- obj@meta.data$tag
idx_list <- split(seq_along(sample_var), sample_var)
idx_selected <- unlist(lapply(idx_list, function(idx) {
  idx[seq_len(min(n_per_sample, length(idx)))]
}), use.names = FALSE)
```

Distribution of CD4_mag expression.

```{r}
set.seed(123)
sc <- obj@meta.data$CD4_mag[idx_selected]
pd <- pdfCluster(sc, hmult=0.75)
cl <- groups(pd)
names(cl) <- names(sc)
table(cl)
tapply(sc, cl, summary)
dens <- density(sc)
pdfcluster_plots[["CD4"]] <- {
  grid.newpage()
  dens <- density(sc)
  plot(dens, main="CD4_mag", ylim=c(-0.005, max(dens$y)))
  abline(v=tapply(sc, cl, min)[-1], lty=3, lwd=2, col='gray')
  abline(v=tapply(sc, cl, max)[-1], lty=3, lwd=2, col='gray')
  abline(h=c(-0.001, -0.005))
  text(x=sc[cl==1], y=-0.001, col='green3', label='|')
  text(x=sc[cl==2], y=-0.005, col='red3', label='|')
  text(x=sc[cl==3], y=-0.005, col='purple', label='|')
  recordPlot()
}
x <- tapply(sc, cl, min)
x_sorted <- sort(x, na.last = NA)
thresholdCD4 <- x_sorted[2]
rm(sc, pd, cl, dens)
```

Distribution of CD8A_mag expression.

```{r}
set.seed(456)
sc <- obj@meta.data$CD8A_mag[idx_selected]
pd <- pdfCluster(sc, hmult=0.75)
cl <- groups(pd)
names(cl) <- names(sc)
table(cl)
tapply(sc, cl, summary)
dens <- density(sc)
pdfcluster_plots[["CD8A"]] <- {
  grid.newpage()
  dens <- density(sc)
  plot(dens, main="CD8A_mag", ylim=c(-0.005, max(dens$y)))
  abline(v=tapply(sc, cl, min)[-1], lty=3, lwd=2, col='gray')
  abline(v=tapply(sc, cl, max)[-1], lty=3, lwd=2, col='gray')
  abline(h=c(-0.001, -0.005))
  text(x=sc[cl==1], y=-0.001, col='green3', label='|')
  text(x=sc[cl==2], y=-0.005, col='red3', label='|')
  text(x=sc[cl==3], y=-0.005, col='purple', label='|')
  recordPlot()
}
x <- tapply(sc, cl, min)
x_sorted <- sort(x, na.last = NA)
thresholdCD8A <- x_sorted[2]
rm(sc, pd, cl, dens)
```

Distribution of CD8B_mag expression.

```{r}
set.seed(789)
sc <- obj@meta.data$CD8B_mag[idx_selected]
pd <- pdfCluster(sc, hmult=0.75)
cl <- groups(pd)
names(cl) <- names(sc)
table(cl)
tapply(sc, cl, summary)
dens <- density(sc)
pdfcluster_plots[["CD8B"]] <- {
  grid.newpage()
  dens <- density(sc)
  plot(dens, main="CD8B_mag", ylim=c(-0.005, max(dens$y)))
  abline(v=tapply(sc, cl, min)[-1], lty=3, lwd=2, col='gray')
  abline(v=tapply(sc, cl, max)[-1], lty=3, lwd=2, col='gray')
  abline(h=c(-0.001, -0.005))
  text(x=sc[cl==1], y=-0.001, col='green3', label='|')
  text(x=sc[cl==2], y=-0.005, col='red3', label='|')
  text(x=sc[cl==3], y=-0.005, col='purple', label='|')
  recordPlot()
}
x <- tapply(sc, cl, min)
x_sorted <- sort(x, na.last = NA)
thresholdCD8B <- x_sorted[2]
rm(sc, pd, cl, dens)
```

Internal control: The consistency of pdfCluster-derived thresholds across samples was assessed to ensure robust discrimination of CD4+ and CD8+ T cells.

```{r, warning=FALSE}
md <- data.frame(CD8A = obj@meta.data$CD8A_mag,
                CD8B = obj@meta.data$CD8B_mag,
                CD4 = obj@meta.data$CD4_mag,
                CD39 = obj@meta.data$ENTPD1_mag,
                PD1 = obj@meta.data$PDCD1_mag,
                clusters = obj@meta.data$seurat_clusters
                )
g1 <- ggplot(md, aes(x = CD8A, y = CD4)) +
  geom_point(
    aes(color = CD8B),
    alpha = 0.3,
    size = 0.8
  ) +
  geom_density_2d(
    aes(group = clusters),
    color = "black",
    linewidth = 0.3,
    alpha = 0.6
  ) +
  scale_color_viridis_c(option = "H") +
  facet_wrap(~clusters) +
  geom_vline(xintercept = thresholdCD8A, linetype = "dashed", linewidth = 0.3) +
  geom_hline(yintercept = thresholdCD4, linetype = "dashed", linewidth = 0.3) +
  labs(
    x = "CD8A",
    y = "CD4",
    color = "CD8B"
  ) +
  theme_ic
g2 <- ggplot(md, aes(x = CD8B, y = CD4)) +
  geom_point(
    aes(color = CD8A),
    alpha = 0.3,
    size = 0.8
  ) +
  geom_density_2d(
    aes(group = clusters),
    color = "black",
    linewidth = 0.3,
    alpha = 0.6
  ) +
  scale_color_viridis_c(option = "H") +
  facet_wrap(~clusters) +
  geom_vline(xintercept = thresholdCD8B, linetype = "dashed", linewidth = 0.3) +
  geom_hline(yintercept = thresholdCD4, linetype = "dashed", linewidth = 0.3) +
  labs(
    x = "CD8B",
    y = "CD4",
    color = "CD8A"
  ) +
  theme_ic
plots_list[["IC09_pdfCluster_thresholds"]] <- wrap_plots(g1, g2, ncol = 2)
print(plots_list[["IC09_pdfCluster_thresholds"]])
rm(g1, g2, md)
```

Cells were classified as CD4+ or CD8+ based on the derived expression thresholds.

```{r}
# --- CD8+ T cells ---
obj$CD8ind <- rep(FALSE, length(colnames(obj)))
obj$CD8ind[obj$CD8A_mag>thresholdCD8A & obj$CD4_mag<thresholdCD4 & !(obj$percent.rpa<20 & obj$percent.mt > 15)] <- TRUE
# --- CD4+ T cells ---
obj$CD4ind <- rep(FALSE, length(colnames(obj)))
obj$CD4ind[obj$CD8A_mag<thresholdCD8A & obj$CD4_mag>thresholdCD4 & !(obj$percent.rpa<20 & obj$percent.mt > 15)] <- TRUE
# --- Consensus ---
obj$magic.method <- ifelse(obj$CD8ind == TRUE & obj$CD4ind == FALSE, "cd8",
                         ifelse(obj$CD8ind == FALSE & obj$CD4ind == TRUE, "cd4", "unknown"))
```

# annotations

Cell identities were further refined using scGate.
<https://github.com/carmonalab/scGate>

```{r, warning=FALSE}
scGate_models_DB <- get_scGateDB()
models.pbmc <- scGate_models_DB$human$PBMC
obj <- scGate(obj, assay = "SCT", model = models.pbmc)
obj$scGate_multi <- ifelse(is.na(obj$scGate_multi), "unknown", obj$scGate_multi)
rm(scGate_models_DB, models.pbmc)
```

Internal control: Evaluate the consistency of Magic-imputed expression with scGate for annotation of CD4+ and CD8+ T cells.

```{r}
table(obj$magic.method, obj$scGate_multi)
p1 <- DimPlot(obj, group.by = "scGate_multi", split.by = "magic.method") + theme(aspect.ratio = 1) + ggtitle("magic.method_vs_scGate")
```

For the final CD4+/CD8+ annotations, scGate was used as the primary lineage classifier, with MAGIC providing complementary evidence. In cases of disagreement, the scGate annotation was prioritised.

```{r}
obj$celltype <- "unknown"
obj$celltype <- ifelse(obj$magic.method == "cd4" & obj$scGate_multi == "CD4T", "cd4", obj$celltype)
obj$celltype <- ifelse(obj$magic.method == "unknown" & obj$scGate_multi == "CD4T", "cd4", obj$celltype)
obj$celltype <- ifelse(obj$magic.method == "cd8" & obj$scGate_multi == "CD4T", "cd4", obj$celltype)
obj$celltype <- ifelse(obj$magic.method == "cd8" & obj$scGate_multi == "CD8T", "cd8", obj$celltype)
obj$celltype <- ifelse(obj$magic.method == "unknown" & obj$scGate_multi == "CD8T", "cd8", obj$celltype)
p2 <- DimPlot(obj, group.by = "scGate_multi", split.by = "celltype") + theme(aspect.ratio = 1) + ggtitle("consensus")
plots_list[["IC10_celltype"]] <- wrap_plots(p1, p2, nrow = 2)
print(plots_list$IC10_celltype)
rm(p1, p2)
```

# tcr analysis

Plots of TCR frequencies for each tool and combinations:

```{r}
# --- Save plots and tables ---
save.plots <- file.path(datapath, sample, folder_data, "plots")
dir.create(save.plots, recursive = F)
save.freqs <- file.path(datapath, sample, folder_data, "frequencies")
dir.create(save.freqs, recursive = F)
# --- 1. Seurat-derived TCRs ---
plot_tcr_frequencies(obj@meta.data %>% filter(celltype == "cd8"),
                     y_log10 = TRUE,
                     cdr3_col = "cdr3",
                     v_gene_col = "v_gene",
                     j_gene_col = "j_gene",
                     chains_col = "chains",
                     save_plot = NULL,
                     save_table = NULL,
                     save_shared_table = NULL
                     )
# --- 2. Combined TCRs ---
plot_tcr_frequencies(obj@meta.data %>% filter(celltype == "cd8"),
                     y_log10 = TRUE,
                     cdr3_col = "cdr3_combi",
                     v_gene_col = "v_gene_combi",
                     j_gene_col = "j_gene_combi",
                     chains_col = "chains_combi",
                     save_plot = NULL,
                     save_table = NULL,
                     save_shared_table = NULL
                     )
# --- 3. Infusion product (IP) vs Blood (PD1hiCD39+) TCRs ---
df <- obj@meta.data %>% filter(celltype == "cd8") %>% dplyr::mutate(tag = ifelse(grepl("Sample1|Sample3", tag), "IP", "PD1hiCD39+"))
plot_tcr_frequencies(meta = df,
                     samples = c("IP", "PD1hiCD39+"),
                     y_log10 = TRUE,
                     cdr3_col = "cdr3_combi",
                     v_gene_col = "v_gene_combi",
                     j_gene_col = "j_gene_combi",
                     chains_col = "chains_combi",
                     save_plot = NULL,
                     save_table = NULL,
                     save_shared_table = NULL
                     )
```

Analyses were performed twice to capture either 70% or 100% of cumulative clonotype frequency.

```{r}
sample_names <- c(
  Sample1 = "5_TIL-15_IP-TIL_4-1BB+_NeoAg_Pool03",
  Sample2 = "6_TIL-15_CD8+_PD-1hiCD39+_PBL_4-1BB+_NeoAg_Pool01",
  Sample3 = "7_TIL-15_IP-TIL_4-1BB+_NeoAg_Pool04",
  Sample4 = "8_TIL-15_CD8+_PD-1hiCD39+_PBL_4-1BB+_NeoAg_Pool02"
)
if (!exists("freq_tables") || !is.list(freq_tables)) freq_tables <- list()
if (!exists("topx_lists") || !is.list(topx_lists)) topx_lists <- list()
max_clonotypes <- 5000
legend_max <- 25
for (coverage_target in c(70, 100)) {
  for (i in names(sample_names)) {
    df <- obj@meta.data %>%
      dplyr::filter(celltype == "cd8") %>%
      dplyr::filter(sample == i) %>%
      dplyr::mutate(
        cdr3_vj = paste(cdr3_combi, v_gene_combi, j_gene_combi, sep = "_"),
        cdr3_vj = sapply(cdr3_vj, function(x) {
          if (is.null(x) || all(is.na(x))) return("no_tcr")
          if (is.list(x)) x <- paste(unlist(x), collapse = ";")
          x
        }),
        cdr3_vj = as.character(cdr3_vj)
      )
    df$cdr3_vj[is.na(df$cdr3_vj) | df$cdr3_vj == "NA_NA_NA"] <- "no_tcr"
    sample_label <- sample_names[[i]]
    freq_table <- df %>%
      dplyr::filter(cdr3_vj != "no_tcr") %>%
      dplyr::group_by(sample, cdr3_vj) %>%
      dplyr::summarise(
        n = dplyr::n(),
        chains_combi = dplyr::first(na.omit(chains_combi)),
        cdr3_combi = dplyr::first(na.omit(cdr3_combi)),
        v_gene_combi = dplyr::first(na.omit(v_gene_combi)),
        d_gene = dplyr::first(na.omit(d_gene)),
        j_gene_combi = dplyr::first(na.omit(j_gene)),
        c_gene = dplyr::first(na.omit(c_gene)),
        reads = dplyr::first(na.omit(reads)),
        umis = dplyr::first(na.omit(umis)),
        productive = dplyr::first(na.omit(productive)),
        full_length = dplyr::first(na.omit(full_length)),
        paired = dplyr::first(na.omit(paired)),
        .groups = "drop"
      ) %>%
      dplyr::mutate(freq = n / sum(n) * 100) %>%
      dplyr::arrange(dplyr::desc(freq)) %>%
      dplyr::select(
        cdr3_vj, freq, n,
        chains_combi, cdr3_combi, v_gene_combi, d_gene, j_gene_combi, c_gene,
        reads, umis, productive, full_length, paired
      )
    topx_tbl <- freq_table %>%
      dplyr::arrange(dplyr::desc(freq)) %>%
      dplyr::mutate(cumfreq = cumsum(freq))
    idx <- which(topx_tbl$cumfreq >= coverage_target)[1]
    if (is.na(idx)) idx <- nrow(topx_tbl)
    idx <- min(idx, max_clonotypes)
    topx_tbl <- topx_tbl %>% dplyr::slice(seq_len(idx)) %>% dplyr::select(-cumfreq)
    freq_tables[[paste0(sample_label, "_full")]] <- freq_table
    freq_tables[[paste0(sample_label, "_top", coverage_target)]] <- topx_tbl
    top_keys <- topx_tbl$cdr3_vj
    topx_lists[[paste0(sample_label, "_top", coverage_target)]] <- topx_tbl %>%
      dplyr::mutate(sample_label = sample_label) %>%
      dplyr::select(sample_label, cdr3_vj, freq, n)
    legend_keys <- head(top_keys, legend_max)
    inside_top_hidden_label <- "legend not shown"
    outside_top_label <- paste0("outside top ", coverage_target, "%")
    n_clones <- length(legend_keys)
    base_palette <- grDevices::colorRampPalette(RColorBrewer::brewer.pal(8, "Paired"))
    base_colors <- base_palette(max(11, n_clones))
    color_manual <- stats::setNames(base_colors[seq_len(n_clones)], legend_keys)
    color_manual[inside_top_hidden_label] <- "black"
    color_manual[outside_top_label] <- "gray70"
    cdr3_counts_sample <- freq_table %>%
      dplyr::mutate(
        cdr3_group = dplyr::case_when(
          cdr3_vj %in% legend_keys ~ cdr3_vj,
          cdr3_vj %in% top_keys ~ inside_top_hidden_label,
          TRUE ~ outside_top_label
        )
      ) %>%
      dplyr::group_by(cdr3_group) %>%
      dplyr::summarise(n = sum(n), .groups = "drop") %>%
      dplyr::mutate(freq = n / sum(n) * 100) %>%
      dplyr::arrange(dplyr::desc(freq))
    plot_levels <- c(
      legend_keys[legend_keys %in% cdr3_counts_sample$cdr3_group],
      inside_top_hidden_label,
      outside_top_label
    )
    plot_levels <- unique(plot_levels[plot_levels %in% cdr3_counts_sample$cdr3_group])
    cdr3_counts_sample <- cdr3_counts_sample %>%
      dplyr::mutate(
        cdr3_group = factor(cdr3_group, levels = plot_levels)
      )
    pie_sample <- ggplot2::ggplot(
      cdr3_counts_sample,
      ggplot2::aes(x = "", y = freq, fill = cdr3_group)
    ) +
      ggplot2::geom_col(color = "black", linewidth = 0.1) +
      ggplot2::coord_polar(theta = "y", direction = 1) +
      ggplot2::scale_fill_manual(values = color_manual, drop = FALSE) +
      ggplot2::labs(
        title = sample_label,
        fill = paste0("CDR3_VJ (top ", coverage_target, "%; max ", max_clonotypes, ")")
      ) +
      ggplot2::theme_void() +
      ggplot2::theme(
        legend.position = "right",
        plot.title = element_text(
          hjust = 0, margin = margin(l=40),
          size = 10
          ),
        plot.margin = margin(t = 10, r = 5, b = 5, l = 5)
    )
    plots_list[[paste0(sample_label, "_top", coverage_target)]] <- pie_sample
  }
  shared_file <- file.path(save.freqs, "shared_ip_blood_cellranger+mixcr_shared.tsv")
  if (file.exists(shared_file) && file.access(shared_file, 4) == 0) {
    shared_IP_blood <- data.table::fread(shared_file) %>%
      dplyr::select(combo, full_tcr) %>%
      dplyr::rename(sample_label = combo, cdr3_vj = full_tcr)
    } else {
      shared_IP_blood <- tibble::tibble(
        sample_label = character(),
        cdr3_vj = character()
      )
      }
  topx_report <- dplyr::bind_rows(
    topx_lists[grepl(paste0("_top", coverage_target, "$"), names(topx_lists))]
  ) %>%
    dplyr::select(sample_label, cdr3_vj)
  topx_report <- dplyr::bind_rows(topx_report, shared_IP_blood) %>%
    dplyr::group_by(cdr3_vj) %>%
    dplyr::summarise(
      sample_label = paste(sort(unique(stats::na.omit(sample_label))), collapse = " | "),
      .groups = "drop"
    )
  write.table(
    topx_report,
    file.path(save.freqs, paste0("TCR_interrogated_top", coverage_target, ".tsv")),
    sep = "\t",
    dec = ",",
    row.names = FALSE,
    quote = FALSE
  )
}
```

# save

```{r}
# --- Save plots ---
dir.create(file.path(datapath, sample, folder_data, "plots"), recursive = FALSE)
for (i in names(plots_list)) {
  outfile <- file.path(datapath, sample, folder_data, "plots", paste0(i, ".pdf"))
  tryCatch({
    pdf(file = outfile, width = 297 / 15, height = 210 / 15)
    print(plots_list[[i]])
    dev.off()
  }, error = function(e) {
    message("Error in plot: ", i)
    message("  ", e$message)
    if (dev.cur() != 1) {
      try(dev.off(), silent = TRUE)
    }
  })
}
# --- Save frequency tables ---
dir.create(file.path(datapath, sample, folder_data, "frequencies"), recursive = F)
for (i in names(freq_tables)) {
  write.table(freq_tables[[i]], file.path(datapath, sample, folder_data, "frequencies", paste0(i, "_cdr3_vj_frequencies.tsv")), sep = "\t", dec = ",", row.names = FALSE, quote = FALSE)
}
# --- Save Seurat object ---
saveRDS(obj, file.path(datapath, sample, folder_data, "seurat/obj_filt.rds"))
```