---
title: "Identification of TCRs-of-interest in peripheral blood"
author: "Endika Prieto <endikaprieto@vhio.net>"
date: "2026-04-24"
---

# setup

```{r, warning=FALSE, message=FALSE}
# --- Load libraries and custom functions ---
set.seed(123)
source("/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/pbl/single_cell/sc_pipeline/scripts/TIL15/setup-m2.R")
# --- Define paths and sample identifiers ---
datapath <- "/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/pbl/single_cell/sc_pipeline/patients"
sample <- "TIL15"
folder_data <- "ITAG_AGG_S01"
# --- Define data loading path ---
datapath_tcrb <- "/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/collabs/Andrea Garcia VHIO/CancerDiscovery_2026/SEQTR/ISF_Rep_2026_00912"
datapath_tcr_validation <- "/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/pbl/single_cell/sc_pipeline/patients/TIL15/ITAG_AGG_S01/outs/validated_tcrs.txt"
# --- Initialize lists for plot storage ---
plots_list <- list()
pdfcluster_plots <- list()
```

# load

```{r}
obj <- readRDS(file.path(datapath, sample, folder_data, "seurat/obj_filt.rds"))
```

# annot

Annotate TCRs identified in the NeoAg pool (top 100% and top 70%) by first running ITAG_TIL15_S02.Rmd.

```{r}
obj@meta.data$cdr3_vj <- ifelse(
  is.na(obj$cdr3_combi) | obj$cdr3_combi == "" | obj$cdr3_combi == "no_tcr",
  "no_tcr",
  paste0(obj$cdr3_combi, "_", obj$v_gene_combi, "_", obj$j_gene_combi)
)
tcr_files <- c(
  pool = "TCR_interrogated_top100.tsv",
  pool_top70 = "TCR_interrogated_top70.tsv"
)
for (anno_col in names(tcr_files)) {
  tcrs_of_interest <- read.table(
    file.path(datapath, sample, "ITAG_AGG_S02", "frequencies", tcr_files[[anno_col]]),
    header = TRUE,
    sep = "\t",
    stringsAsFactors = FALSE
  )
  idx <- match(obj@meta.data$cdr3_vj, tcrs_of_interest$cdr3_vj)
  obj@meta.data[[anno_col]] <- tcrs_of_interest$sample_label[idx]
}
```

Annotate cell compartments based on the tag and cell type information.

Note: Sorted populations were supplemented with cells from the antagonistic compartment (i.e., opposite lineage and/or CD39 status) to meet minimum cell requirements for downstream analyses.

```{r}
obj$tag.compartment <- dplyr::case_when(
  obj$tag == "Sample1" & obj$celltype == "cd8" ~ "CD8+CD39+",
  obj$tag == "Sample1" & obj$celltype == "cd4" ~ "CD4+CD39-",
  obj$tag == "Sample2" & obj$celltype == "cd4" ~ "CD4+CD39+",
  obj$tag == "Sample2" & obj$celltype == "cd8" ~ "CD8+CD39-",
  obj$tag == "Sample3" & obj$celltype == "cd8" ~ "CD8+Bulk",
  obj$tag == "Sample3" & obj$celltype == "cd4" ~ "CD4+Bulk",
  obj$tag == "Sample4" & obj$celltype == "cd4" ~ "CD4+Bulk",
  TRUE ~ NA_character_
)
table(obj$tag.compartment)
```

PDCD1 levels for each compartment.

```{r}
# --- CD4+Bulk ---
cd4_bulk <- subset(obj, subset = tag.compartment == "CD4+Bulk")
idx_selected <- select_representative_indices(cd4_bulk@meta.data, n_total = 1500, tag_col = "tag")
sc <- cd4_bulk@meta.data$PDCD1_mag[idx_selected]
pd <- pdfCluster(sc, hmult = 0.75)
summary(pd)
cl <- pdfCluster::groups(pd)
names(cl) <- names(sc)
table(cl)
tapply(sc, cl, summary)
pdfcluster_plots[["PDCD1_CD4_Bulk"]] <- {
  grid.newpage()
  dens <- density(sc)
  plot(dens, main = "PDCD1_CD4_Bulk", ylim = c(-0.005, max(dens$y)))
  abline(v = tapply(sc, cl, min)[-1], lty = 3, lwd = 2, col = "gray")
  abline(v = tapply(sc, cl, max)[-1], lty = 3, lwd = 2, col = "gray")
  abline(h = c(-0.001, -0.005))
  text(x = sc[cl == 1], y = -0.001, col = "green3", labels = "|")
  text(x = sc[cl == 2], y = -0.005, col = "red3", labels = "|")
  if (sum(cl == 3) > 0) text(x = sc[cl == 3], y = -0.005, col = "purple", labels = "|")
  recordPlot()
}
cluster_vals <- split(sc, cl)
cluster_medians <- sapply(cluster_vals, median)
cluster_order <- order(cluster_medians)
threshold_PDCD1_cd4_bulk <- if (length(cluster_order) >= 2) {
  min(cluster_vals[[cluster_order[2]]])
} else {
  NA_real_
}
rm(sc, pd, cl, cluster_vals, cluster_medians, cluster_order)
```

```{r}
# --- CD4+CD39+ ---
cd4_cd39 <- subset(obj, subset = tag.compartment == "CD4+CD39+")
idx_selected <- select_representative_indices(cd4_cd39@meta.data, n_total = 1500, tag_col = "tag")
sc <- cd4_cd39@meta.data$PDCD1_mag[idx_selected]
pd <- pdfCluster(sc, hmult = 0.75)
summary(pd)
cl <- pdfCluster::groups(pd)
names(cl) <- names(sc)
table(cl)
tapply(sc, cl, summary)
pdfcluster_plots[["PDCD1_CD4_CD39"]] <- {
  grid.newpage()
  dens <- density(sc)
  plot(dens, main = "PDCD1_CD4_CD39", ylim = c(-0.005, max(dens$y)))
  abline(v = tapply(sc, cl, min)[-1], lty = 3, lwd = 2, col = "gray")
  abline(v = tapply(sc, cl, max)[-1], lty = 3, lwd = 2, col = "gray")
  abline(h = c(-0.001, -0.005))
  text(x = sc[cl == 1], y = -0.001, col = "green3", labels = "|")
  text(x = sc[cl == 2], y = -0.005, col = "red3", labels = "|")
  if (sum(cl == 3) > 0) text(x = sc[cl == 3], y = -0.005, col = "purple", labels = "|")
  recordPlot()
}
cluster_vals <- split(sc, cl)
cluster_medians <- sapply(cluster_vals, median)
cluster_order <- order(cluster_medians)
threshold_PDCD1_cd4_cd39 <- if (length(cluster_order) >= 2) {
  min(cluster_vals[[cluster_order[2]]])
} else {
  NA_real_
}
rm(sc, pd, cl, cluster_vals, cluster_medians, cluster_order)
```

```{r}
# --- CD4+CD39neg ---
cd4_cd39neg <- subset(obj, subset = tag.compartment == "CD4+CD39-")
idx_selected <- select_representative_indices(cd4_cd39neg@meta.data, n_total = 1500, tag_col = "tag")
sc <- cd4_cd39neg@meta.data$PDCD1_mag[idx_selected]
pd <- pdfCluster(sc, hmult = 0.75)
summary(pd)
cl <- pdfCluster::groups(pd)
names(cl) <- names(sc)
table(cl)
tapply(sc, cl, summary)
pdfcluster_plots[["PDCD1_CD4_CD39neg"]] <- {
  grid.newpage()
  dens <- density(sc)
  plot(dens, main = "PDCD1_CD4_CD39neg", ylim = c(-0.005, max(dens$y)))
  abline(v = tapply(sc, cl, min)[-1], lty = 3, lwd = 2, col = "gray")
  abline(v = tapply(sc, cl, max)[-1], lty = 3, lwd = 2, col = "gray")
  abline(h = c(-0.001, -0.005))
  text(x = sc[cl == 1], y = -0.001, col = "green3", labels = "|")
  text(x = sc[cl == 2], y = -0.005, col = "red3", labels = "|")
  if (sum(cl == 3) > 0) text(x = sc[cl == 3], y = -0.005, col = "purple", labels = "|")
  recordPlot()
}
cluster_vals <- split(sc, cl)
cluster_medians <- sapply(cluster_vals, median)
cluster_order <- order(cluster_medians)
threshold_PDCD1_cd4_cd39neg <- if (length(cluster_order) >= 2) {
  min(cluster_vals[[cluster_order[2]]])
} else {
  NA_real_
}
rm(sc, pd, cl, cluster_vals, cluster_medians, cluster_order)
```

```{r}
# --- CD8+Bulk ---
cd8_bulk <- subset(obj, subset = tag.compartment == "CD8+Bulk")
idx_selected <- select_representative_indices(cd8_bulk@meta.data, n_total = 1000, tag_col = "tag")
sc <- cd8_bulk@meta.data$PDCD1_mag[idx_selected]
pd <- pdfCluster(sc, hmult = 0.75)
summary(pd)
cl <- pdfCluster::groups(pd)
names(cl) <- names(sc)
table(cl)
tapply(sc, cl, summary)
pdfcluster_plots[["PDCD1_CD8_Bulk"]] <- {
  grid.newpage()
  dens <- density(sc)
  plot(dens, main = "PDCD1_CD8_Bulk", ylim = c(-0.005, max(dens$y)))
  abline(v = tapply(sc, cl, min)[-1], lty = 3, lwd = 2, col = "gray")
  abline(v = tapply(sc, cl, max)[-1], lty = 3, lwd = 2, col = "gray")
  abline(h = c(-0.001, -0.005))
  text(x = sc[cl == 1], y = -0.001, col = "green3", labels = "|")
  text(x = sc[cl == 2], y = -0.005, col = "red3", labels = "|")
  if (sum(cl == 3) > 0) text(x = sc[cl == 3], y = -0.005, col = "purple", labels = "|")
  recordPlot()
}
cluster_vals <- split(sc, cl)
cluster_medians <- sapply(cluster_vals, median)
cluster_order <- order(cluster_medians)
threshold_PDCD1_cd8_bulk <- if (length(cluster_order) >= 2) {
  min(cluster_vals[[cluster_order[2]]])
} else {
  NA_real_
}
rm(sc, pd, cl, cluster_vals, cluster_medians, cluster_order)
```

```{r}
# --- CD8+CD39+ ---
cd8_cd39 <- subset(obj, subset = tag.compartment == "CD8+CD39+")
idx_selected <- select_representative_indices(cd8_cd39@meta.data, n_total = 1500, tag_col = "tag")
sc <- cd8_cd39@meta.data$PDCD1_mag[idx_selected]
pd <- pdfCluster(sc, hmult = 0.75)
summary(pd)
cl <- pdfCluster::groups(pd)
names(cl) <- names(sc)
table(cl)
tapply(sc, cl, summary)
pdfcluster_plots[["PDCD1_CD8_CD39"]] <- {
  grid.newpage()
  dens <- density(sc)
  plot(dens, main = "PDCD1_CD8_CD39", ylim = c(-0.005, max(dens$y)))
  abline(v = tapply(sc, cl, min)[-1], lty = 3, lwd = 2, col = "gray")
  abline(v = tapply(sc, cl, max)[-1], lty = 3, lwd = 2, col = "gray")
  abline(h = c(-0.001, -0.005))
  text(x = sc[cl == 1], y = -0.001, col = "green3", labels = "|")
  text(x = sc[cl == 2], y = -0.005, col = "red3", labels = "|")
  if (sum(cl == 3) > 0) text(x = sc[cl == 3], y = -0.005, col = "purple", labels = "|")
  recordPlot()
}
cluster_vals <- split(sc, cl)
cluster_medians <- sapply(cluster_vals, median)
cluster_order <- order(cluster_medians)
threshold_PDCD1_cd8_cd39 <- if (length(cluster_order) >= 2) {
  min(cluster_vals[[cluster_order[2]]])
} else {
  NA_real_
}
rm(sc, pd, cl, cluster_vals, cluster_medians, cluster_order)
```

```{r}
# --- CD8+CD39neg ---
cd8_cd39neg <- subset(obj, subset = tag.compartment == "CD8+CD39-")
idx_selected <- select_representative_indices(cd8_cd39neg@meta.data, n_total = 1500, tag_col = "tag")
sc <- cd8_cd39neg@meta.data$PDCD1_mag[idx_selected]
pd <- pdfCluster(sc, hmult = 0.75)
summary(pd)
cl <- pdfCluster::groups(pd)
names(cl) <- names(sc)
table(cl)
tapply(sc, cl, summary)
pdfcluster_plots[["PDCD1_CD8_CD39neg"]] <- {
  grid.newpage()
  dens <- density(sc)
  plot(dens, main = "PDCD1_CD8_CD39neg", ylim = c(-0.005, max(dens$y)))
  abline(v = tapply(sc, cl, min)[-1], lty = 3, lwd = 2, col = "gray")
  abline(v = tapply(sc, cl, max)[-1], lty = 3, lwd = 2, col = "gray")
  abline(h = c(-0.001, -0.005))
  text(x = sc[cl == 1], y = -0.001, col = "green3", labels = "|")
  text(x = sc[cl == 2], y = -0.005, col = "red3", labels = "|")
  if (sum(cl == 3) > 0) text(x = sc[cl == 3], y = -0.005, col = "purple", labels = "|")
  recordPlot()
}
cluster_vals <- split(sc, cl)
cluster_medians <- sapply(cluster_vals, median)
cluster_order <- order(cluster_medians)
threshold_PDCD1_cd8_cd39neg <- if (length(cluster_order) >= 2) {
  min(cluster_vals[[cluster_order[2]]])
} else {
  NA_real_
}
rm(sc, pd, cl, cluster_vals, cluster_medians, cluster_order)
```

# prior

Aggregate and validate TCR clonotypes across CD4/CD8 compartments, summarizing cell counts per clonotype and annotating NeoAg-derived TCRs (top 100% and top 70%).

```{r}
sample_names <- c("CD4+Bulk","CD4+CD39-","CD4+CD39+","CD8+Bulk","CD8+CD39-","CD8+CD39+")
captured_tables <- set_names(sample_names) %>%
  purrr::map(\(grp)
    obj@meta.data |>
      dplyr::filter(tag.compartment == grp) |>
      dplyr::mutate(
        pool         = as.character(pool),
        chains_combi = as.character(chains_combi),
        cdr3_combi   = as.character(cdr3_combi),
        v_gene_combi = as.character(v_gene_combi),
        j_gene_combi = as.character(j_gene_combi),
        cdr3_vj      = as.character(cdr3_vj)
      ) |>
      dplyr::count(
        pool, chains_combi, cdr3_combi, v_gene_combi, j_gene_combi, cdr3_vj,
        name = "n_cells", .drop = FALSE
      ) |>
      dplyr::arrange(pool, dplyr::desc(n_cells))
  )
src_map <- c("CD8+Bulk" = "Bulk", "CD8+CD39+" = "CD39+", "CD8+CD39-" = "CD39-")
abc <- imap_dfr(src_map, \(src, grp)
  captured_tables[[grp]] |>
    filter(!is.na(pool)) |>
    mutate(source = src)
)
tcrs_of_interest_top70 <- read.table(
  file.path(datapath, sample, folder_data, "frequencies", "TCR_interrogated_top70.tsv"),
  header = TRUE, sep = "\t", stringsAsFactors = FALSE
)
validation <- abc %>%
  group_by(pool, chains_combi, cdr3_combi, v_gene_combi, j_gene_combi, cdr3_vj) %>%
  summarise(
    n_cells = paste(n_cells[order(source)], collapse = " | "),
    source  = paste(source[order(source)], collapse = " | "),
    .groups = "drop"
  ) %>%
  arrange(pool, desc(as.numeric(sub(" \\|.*$", "", n_cells)))) %>%
  mutate(
    top100 = "Y",
    top70  = if_else(cdr3_vj %in% tcrs_of_interest_top70$cdr3_vj, "Y", "N")
  )
```

Controls: Highly expanded CD8+ T cells.

```{r}
md_f <- obj@meta.data %>%
  dplyr::filter(
    tag.compartment %in% c("CD8+Bulk", "CD8+CD39+", "CD8+CD39-"),
    is.na(pool)
  ) %>%
  dplyr::mutate(
    source = dplyr::recode(tag.compartment,
      "CD8+Bulk"  = "Bulk",
      "CD8+CD39+" = "CD39+",
      "CD8+CD39-" = "CD39-"
    )
  )
selected_clonos <- md_f %>%
  dplyr::count(
    source, chains_combi, cdr3_combi, v_gene_combi, j_gene_combi, cdr3_vj,
    name = "n_cells"
  ) %>%
  dplyr::group_by(source) %>%
  dplyr::slice_max(n_cells, n = 3, with_ties = TRUE) %>%
  dplyr::ungroup() %>%
  dplyr::distinct(chains_combi, cdr3_combi, v_gene_combi, j_gene_combi, cdr3_vj)
expanded_merged <- md_f %>%
  dplyr::semi_join(
    selected_clonos,
    by = c("chains_combi","cdr3_combi","v_gene_combi","j_gene_combi","cdr3_vj")
  ) %>%
  dplyr::count(
    source, chains_combi, cdr3_combi, v_gene_combi, j_gene_combi, cdr3_vj,
    name = "n_cells"
  ) %>%
  tidyr::complete(
    source = c("Bulk","CD39+","CD39-"),
    tidyr::nesting(chains_combi, cdr3_combi, v_gene_combi, j_gene_combi, cdr3_vj),
    fill = list(n_cells = 0)
  ) %>%
  dplyr::group_by(chains_combi, cdr3_combi, v_gene_combi, j_gene_combi, cdr3_vj) %>%
  dplyr::summarise(
    source  = paste(source, collapse = " | "),
    n_cells = paste(n_cells, collapse = " | "),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    pool   = "Highly expanded controls",
    top100 = "N",
    top70  = "N",
    n_cells = as.character(n_cells)
  ) %>%
  dplyr::select(
    pool, chains_combi, cdr3_combi, v_gene_combi, j_gene_combi,
    cdr3_vj, n_cells, source, top100, top70
  )
validation <- dplyr::bind_rows(validation, expanded_merged)
```

Controls: CD8+ T cells exhibiting high or low PDCD1 expression.

```{r}
res_hi_cd39pos <- make_pdcd1_controls(obj@meta.data, "CD8+CD39+", "Controls: CD8+CD39+_PDCD1hi", n_pick = 25, direction = "high")
res_hi_bulk    <- make_pdcd1_controls(obj@meta.data, "CD8+Bulk",  "Controls: CD8+Bulk_PDCD1hi",   n_pick = 10, direction = "high")
validation <- bind_rows(validation, res_hi_cd39pos$table, res_hi_bulk$table)
res_lo_cd39pos <- make_pdcd1_controls(obj@meta.data, "CD8+CD39+", "Controls: CD8+CD39+_PDCD1lo", n_pick = 5, direction = "low")
validation <- bind_rows(validation, res_lo_cd39pos$table)
wrap_plots(res_hi_cd39pos$plot, res_hi_bulk$plot, res_lo_cd39pos$plot, nrow = 1)
```

Annotate the frequency of these TCRβ clonotypes in the tumor sample.

```{r}
AGa0002 <- fread(file.path(datapath_tcrb, "AGa0002-2026-00912_TCRb_repertoire_corrected.txt")) %>%
  dplyr::filter(Frame == "IN") %>%
  dplyr::mutate(freq = Count / sum(Count))
AGa0002 <- AGa0002 %>%
  mutate(
    TRBV = TRBV %>%
      str_replace("^hTRBV", "TRBV") %>%
      str_replace("^TRBV0([1-9])", "TRBV\\1") %>%
      str_replace("-1$", "")
  ) %>%
  mutate(
    TRBJ = TRBJ %>%
      str_replace("^hTRBJ", "TRBJ") %>%
      str_replace("^TRBJ0([1-9])", "TRBJ\\1") %>%
      str_replace("-1$", "")) %>%
  mutate(
    CDR3_aaseq = str_replace(CDR3_aaseq, ".$", "")
  ) %>%
  mutate(
    TCRb = paste0(CDR3_aaseq, "_", TRBV, "_", TRBJ)
  ) %>%
  select(TCRb, freq) %>%
  arrange(desc(freq))
AGa0002_collapsed <- AGa0002 %>%
  group_by(TCRb) %>%
  summarise(
    freq = sum(freq, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(freq))
validation <- validation %>%
  mutate(
    TCRb = pmap_chr(
      list(chains_combi, cdr3_combi, v_gene_combi, j_gene_combi),
      \(chains, cdr3, v, j) {
        ch  <- str_split(chains, ";", simplify = TRUE)
        idx <- which(ch == "TRB")[1]
        if (length(idx) == 0 || is.na(idx)) return(NA_character_)
        cdr3_b <- str_split(cdr3, ";", simplify = TRUE)[1, idx]
        v_b    <- str_split(v,    ";", simplify = TRUE)[1, idx]
        j_b    <- str_split(j,    ";", simplify = TRUE)[1, idx]
        v_b <- str_remove(v_b, "-1$")
        j_b <- str_remove(j_b, "-1$")

        str_c(cdr3_b, v_b, j_b, sep = "_")
      }
    )
  )
validation <- validation %>%
  left_join(AGa0002_collapsed, by = "TCRb") %>%
  filter(chains_combi %in% c("TRA;TRA;TRB", "TRA;TRB")) %>%
  select(-cdr3_vj) %>%
  select(chains_combi, cdr3_combi, v_gene_combi, j_gene_combi, n_cells, pool, source, top100, top70, TCRb, freq)
```

Collapse and save the table.

```{r}
validation <- validation %>%
  dplyr::group_by(cdr3_combi, v_gene_combi, j_gene_combi) %>%
  dplyr::summarise(
    chains_combi = validation_collapse_col(chains_combi),
    n_cells      = validation_collapse_col(n_cells),
    pool         = validation_collapse_col(pool),
    source       = validation_collapse_col(source),
    top100       = validation_collapse_col(top100),
    top70        = validation_collapse_col(top70),
    TCRb         = validation_collapse_col(TCRb),
    freq         = validation_collapse_col(freq),
    .groups = "drop"
  ) %>%
  dplyr::select(
    chains_combi, cdr3_combi, v_gene_combi, j_gene_combi,
    n_cells, pool, source, top100, top70, TCRb, freq
  )
# --- Save a table with TCR candidates for validation ---
timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
outfile <- file.path(
  datapath, sample, folder_data, "validation",
  paste0(timestamp, "_TCR_candidates.tsv")
)
write.table(
  validation,
  file = outfile,
  sep = "\t",
  dec = ",",
  row.names = FALSE,
  quote = FALSE
)
```

⚠️ Note: At this stage, TCRs of interest and control TCRs were prioritized for individual validation.

```{r}
annot <- fread(datapath_tcr_validation) %>%
  select(cdr3_combi, v_gene_combi, j_gene_combi, Pools, TCR.ID, Reactive) %>% distinct() %>% as.data.frame()
obj@meta.data <- obj@meta.data %>%
  rownames_to_column("cell") %>%
  left_join(annot, by = c("cdr3_combi", "v_gene_combi", "j_gene_combi")) %>%
  column_to_rownames("cell")
table(obj$TCR.ID)
# Palette
set.seed(123)
tcr_ids <- sort(unique(na.omit(annot$TCR.ID)))
tcr_colors <- setNames(
  Polychrome::createPalette(
    length(tcr_ids),
    seedcolors = c("#5DA5DA","#F15854","#60BD68","#B276B2")
  ),
  tcr_ids
)
```

# bulk

Add TCRβ clonotype frequencies from the tumor, CD8_PD1hi_CD39pos, Infusion_Product, and total PBMC samples for this patient.

```{r}
bulk.data <- list()
samples <- c("AGa0002", "AGa0007", "AGa0008", "AGa0009")
for (i in samples) {
  data.path <- "/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/collabs/Andrea Garcia VHIO/CancerDiscovery_2026/SEQTR/ISF_Rep_2026_00912/"
  data_i <- fread(file.path(data.path, paste0(i, "-2026-00912_TCRb_repertoire_corrected.txt"))) %>%
    dplyr::filter(Frame == "IN") %>%
    dplyr::mutate(freq = Count / sum(Count)) %>%
    mutate(
      TRBV = TRBV %>%
        str_replace("^hTRBV", "TRBV") %>%
        str_replace("^TRBV0([1-9])", "TRBV\\1")
    ) %>%
    mutate(
      TRBJ = TRBJ %>%
        str_replace("^hTRBJ", "TRBJ") %>%
        str_replace("^TRBJ0([1-9])", "TRBJ\\1")
    ) %>%
    mutate(
      CDR3_aaseq = str_replace(CDR3_aaseq, ".$", "")
    ) %>%
    mutate(
      TCRb = paste0(CDR3_aaseq, "_", TRBV, "_", TRBJ)
    ) %>%
    select(TCRb, freq) %>%
    arrange(desc(freq)) %>%
    group_by(TCRb) %>%
    summarise(
      freq = sum(freq, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(freq))
  bulk.data[[i]] <- data_i
}
bulk.data.wide <- imap_dfr(bulk.data, ~ mutate(.x, sample = .y)) %>%
  distinct(sample, TCRb, .keep_all = TRUE) %>%
  pivot_wider(
    names_from  = sample,
    values_from = freq,
    values_fill = NA_real_
  ) %>%
  rename_with(
    ~ recode(.x,
      AGa0002 = "TIL15_RNA_T",
      AGa0007 = "TIL15_CD8_PD1hi_CD39pos",
      AGa0008 = "TIL_Infusion_Product",
      AGa0009 = "TIL15_PBMCs",
      .default = .x
    )
  )

obj@meta.data <- obj@meta.data %>%
  mutate(
    TCRb = pmap_chr(
      list(chains_combi, cdr3_combi, v_gene_combi, j_gene_combi),
      make_trb_pheno
    )
  )
obj@meta.data <- obj@meta.data %>%
  rownames_to_column(var = "cell") %>%
  left_join(bulk.data.wide, by = "TCRb") %>%
  column_to_rownames(var = "cell")
```

# ranks

Rank CD8+ T cells across compartments (bulk, CD39+, and CD39−) by PDCD1 expression, highlighting NeoAg-derived TCRs and controls.

```{r}
t <- obj@meta.data %>%
  dplyr::select(cdr3_vj, TCR.ID, tag.compartment, Pools, PDCD1_mag, ENTPD1_mag, Reactive)
t2 <- t %>%
  dplyr::mutate(
    tag.compartment = factor(tag.compartment),
    tcr_id_plot = dplyr::case_when(
      is.na(TCR.ID) ~ "NA",
      TRUE ~ as.character(TCR.ID)
    ),
    pool_plot = dplyr::case_when(
      is.na(Pools) ~ "NA",
      TRUE ~ as.character(Pools)
    ),
    reactive_plot = dplyr::case_when(
      Reactive == "Yes" ~ "Yes",
      Reactive == "No" ~ "No",
      TRUE ~ "NA"
    ),
    entpd1_plot = dplyr::case_when(
      is.na(ENTPD1_mag) ~ 0,
      TRUE ~ pmin(pmax(as.numeric(ENTPD1_mag), 0), 1)
    )
  )
pool_levels_raw <- sort(unique(as.character(t2$pool_plot)))
pool_levels <- c(setdiff(pool_levels_raw, "NA"), "NA")
cols_pool <- c(
  "High PDCD1 controls" = "#3C5488",
  "Low PDCD1 controls" = "#5FA8DD",
  "NeoAg_pools" = "#F25856",
  "Highly expanded controls" = "#AD73AD",
  "NA" = "gray85"
)
extra_pool_levels <- setdiff(pool_levels, names(cols_pool))
if (length(extra_pool_levels) > 0) {
  extra_cols <- make_qual_palette(length(extra_pool_levels), palette = "Dark2")
  cols_pool <- c(cols_pool, setNames(extra_cols, extra_pool_levels))
}
cols_pool <- cols_pool[pool_levels]
tcr_levels_raw <- sort(unique(as.character(t2$tcr_id_plot)))
tcr_levels <- c(setdiff(tcr_levels_raw, "NA"), "NA")
tcr_colors_use <- setNames(rep("gray85", length(tcr_levels)), tcr_levels)
tcr_colors_use["NA"] <- "gray85"
tcr_colors_use[names(tcr_colors)] <- ifelse(is.na(tcr_colors), "gray85", unname(tcr_colors))
cols_reactive <- c("Yes" = "red3", "No" = "black", "NA" = "gray85")
tcr_breaks <- names(tcr_colors_use)
pool_breaks <- c(setdiff(pool_levels, "NA"), "NA")
reactive_breaks <- c("Yes", "No", "NA")
tcr_ncol <- max(1, ceiling(length(tcr_breaks) / 5))
pool_ncol <- max(1, ceiling(length(pool_breaks) / 5))
reactive_ncol <- max(1, ceiling(length(reactive_breaks) / 5))
# --- Rank plots ---
p1 <- make_barplot_single(
  t2,
  "CD8+CD39+",
  threshold = threshold_PDCD1_cd8_bulk,
  tcr_colors = tcr_colors_use,
  cols_pool = cols_pool,
  cols_reactive = cols_reactive
)
p2 <- make_barplot_single(
  t2,
  "CD8+Bulk",
  top_n = NULL,
  threshold = threshold_PDCD1_cd8_bulk,
  tcr_colors = tcr_colors_use,
  cols_pool = cols_pool,
  cols_reactive = cols_reactive
)
p2.zoom <- make_barplot_single(
  t2,
  "CD8+Bulk",
  top_n = 300,
  threshold = threshold_PDCD1_cd8_bulk,
  tcr_colors = tcr_colors_use,
  cols_pool = cols_pool,
  cols_reactive = cols_reactive
)
p3 <- make_barplot_single(
  t2,
  "CD8+CD39-",
  top_n = NULL,
  threshold = threshold_PDCD1_cd8_bulk,
  tcr_colors = tcr_colors_use,
  cols_pool = cols_pool,
  cols_reactive = cols_reactive
)
plots_list[["Rank_CD8+"]] <- wrap_plots(p1, p2, p3, nrow = 3, guides = "collect") & theme(legend.position = "right")
plots_list[["Rank_CD8+Bulk_Zoom"]] <- wrap_plots(p2, p2.zoom, nrow = 2, guides = "collect") & theme(legend.position = "right")
```

# cd8

Subset CD8+ T cells.

```{r}
obj.cd8 <- subset(obj, subset = celltype == "cd8")
DefaultAssay(obj.cd8) <- "RNA"
ndim <- 6
res <- 0.75
set.seed(123)
obj.cd8.int <- obj.cd8 %>%
  NormalizeData() %>%
  ScaleData() %>%
  FindVariableFeatures() %>%
  RunPCA(npcs = ndim) %>%
  RunUMAP(dims = 1:ndim, seed.use = 123) %>%
  FindNeighbors(dims = 1:ndim) %>%
  FindClusters(resolution = res, random.seed = 123)
p1 <- DimPlot(obj.cd8.int,
        reduction = "umap",
        group.by = "seurat_clusters",
        label = T,
        repel = F,
        seed = 123,
        label.size = 5) + 
  theme(aspect.ratio = 1)
p2 <- VlnPlot(obj.cd8.int, features = c("TRAC", "CD3D", "CD8A", "CD4"), group.by = "seurat_clusters", pt.size = 0, ncol = 1)
wrap_plots(p1, p2, nrow = 1)
```

Refine clusters.

```{r, warning=FALSE}
t <- data.frame(
  doublets = obj.cd8.int@meta.data$scDblFinder.score,
  logreads = obj.cd8.int@meta.data$logreads,
  percent.mt = obj.cd8.int@meta.data$percent.mt,
  clusters = obj.cd8.int@meta.data$seurat_clusters
  )
g1 <- ggplot(t, aes(x = doublets, y = percent.mt)) + geom_point(aes(color = logreads), alpha = 0.5,size = 0.8) + scale_color_viridis_c(option = "H") + geom_density_2d(aes(color = logreads), alpha = 0.8) + facet_wrap(~clusters) + geom_vline(xintercept = 0.125, lty=2)  + geom_hline(yintercept = 6.5, lty=2) + theme(aspect.ratio = 1) + labs(title = "doublets vs logreads")
print(g1)
```

Filter and re-run dimensionality reduction and clustering.

```{r}
md <- obj.cd8.int@meta.data
cells.keep <- rownames(md)[
  !(md$percent.mt > 7.5 | md$scDblFinder.score > 0.125)
]
# --- Filter low quality cells ---
obj.cd8.int <- subset(
  obj.cd8.int,
  cells = cells.keep
)
obj.cd8.int <- obj.cd8.int %>%
  FindNeighbors(dims = 1:ndim) %>%
  FindClusters(resolution = res) %>%
  RunUMAP(dims = 1:ndim, seed.use = 123)
# --- UMAP plot ---
DimPlot(obj.cd8.int,
        reduction = "umap",
        group.by = "seurat_clusters",
        label = T,
        repel = F,
        seed = 123,
        label.size = 5) + 
  theme(aspect.ratio = 1)
```
Filter cluster 10.

```{r, warning=FALSE, message=FALSE}
obj.cd8.int <- subset(
  obj.cd8.int,
  subset = seurat_clusters != 10
)
obj.cd8.int <- obj.cd8.int %>%
  FindNeighbors(dims = 1:ndim) %>%
  FindClusters(resolution = res) %>%
  RunUMAP(dims = 1:ndim, seed.use = 123)
# --- UMAP plot ---
DimPlot(obj.cd8.int,
        reduction = "umap",
        group.by = "seurat_clusters",
        label = T,
        repel = F,
        seed = 123,
        label.size = 5) + 
  theme(aspect.ratio = 1)
```

Feature plots on UMAP embeddings.

```{r, message=FALSE}
# --- Load magic data ---
magic.data <- readRDS(file.path(datapath, sample, folder_data, "seurat", "magic.rds"))
genes <- c("GZMK", "ENTPD1", "HLA-DRA", "HAVCR2", "TIGIT", "LAG3", "TNFRSF18", "TNFRSF9", "ITGA1", "ITGA2", "ITGAE", "CD45RO", "CXCL13", "TOX", "GZMB", "PRF1", "NKG7",
           "CD27", "CD28", "IL2RA", "CD69", "IFNG",
           "TCF7", "LEF1", "LTB", "KLF2", "SELL","CCR7", "IL7R")
common_cells <- intersect(colnames(obj.cd8.int), colnames(magic.data))
genes_present <- genes[genes %in% rownames(magic.data@assays$RNA$data)]
genes_missing <- setdiff(genes, genes_present)
# --- Plots ---
for (g in genes_present) {
  vals <- magic.data@assays$RNA$data[g, common_cells]
  new_col <- paste0(g, "_mag")
  obj.cd8.int[[new_col]] <- NA_real_
  obj.cd8.int@meta.data[common_cells, new_col] <- as.numeric(vals)
}
p2 <- FeaturePlot(
  obj.cd8.int,
  features = c("PDCD1_mag", "ENTPD1_mag", paste0(genes, "_mag")),
  ncol = 6, pt.size = 0.25
) &
  scale_color_gradientn(
    colours = c("#D9F2E6", "#52B788", "#2C7FB8", "#08306B"),
    oob = scales::squish,
    na.value = "grey80"
  ) &
  theme(aspect.ratio = 1)
plots_list[["Umap_CD8_Features"]] <- p2
```

# pheno

Manual annotation based on cluster pseudobulk analysis.

```{r}
DefaultAssay(obj.cd8.int) <- "RNA"
obj.cd8.int <- NormalizeData(obj.cd8.int)
all_markers <- FindAllMarkers(
  obj.cd8.int,
  group.by = "seurat_clusters",
  logfc.threshold = 0.75,
  min.pct = 0.25,
  only.pos = TRUE
)
top_n_per_cluster <- all_markers %>%
  filter(p_val_adj < 0.05) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), p_val_adj, gene, .by_group = TRUE) %>%
  slice_head(n = 30) %>%
  ungroup()
DefaultAssay(obj.cd8.int) <- "RNA"
Idents(obj.cd8.int) <- "seurat_clusters"
genes_check <- unique(top_n_per_cluster$gene)
avg <- AggregateExpression(
  obj.cd8.int,
  assays = "RNA",
  features = genes_check,
  group.by = "seurat_clusters",
  slot = "data",
  return.seurat = FALSE
)$RNA
avg <- as.matrix(avg)
avg <- avg[apply(avg, 1, function(x) all(is.finite(x))), , drop = FALSE]
avg <- avg[apply(avg, 1, sd, na.rm = TRUE) > 0, , drop = FALSE]
plots_list[["Heatmap_Markers_CD8"]] <- pheatmap::pheatmap(
  t(avg),
  scale = "column",
  clustering_method = "complete",
  fontsize_row = 7,
  fontsize_col = 10,
  cellwidth = 8,
  cellheight = 8,
  main = "AggregateExpression (RNA data) per seurat_cluster"
)
```

Top dysregulated markers in each CD8+ T cell cluster.

```{r}
top_n_per_cluster %>%
  split(.$cluster) %>%
  lapply(\(df) paste(df$gene, collapse = ", ")) %>%
  imap_chr(~ paste0(.y, ": ", .x)) %>%
  cat(sep = "\n")
```

Assign a phenotype or cellular state to each cluster.

```{r}
cluster_labels <- c(
  "0" = "CD8_Activated_PD1pos",
  "1" = "CD8_TEMRA",
  "2" = "CD8_Naive_TCM_like",
  "3" = "CD8_NK_Like",
  "4" = "CD8_Transitional_EffectorMemory",
  "5" = "CD8_EarlyMemory_like",
  "6" = "CD8_Activated",
  "7" = "CD8_ChronicallyActivated_CXCL13pos",
  "8" = "MAIT",
  "9" = "CD8_StemLike_Memory"
)
levels_order <- c(
  "CD8_Naive_TCM_like",
  "CD8_EarlyMemory_like",
  "CD8_StemLike_Memory",
  "CD8_Transitional_EffectorMemory",
  "CD8_Activated",
  "CD8_Activated_PD1pos",
  "CD8_ChronicallyActivated_CXCL13pos",
  "CD8_TEMRA",
  "CD8_NK_Like",
  "MAIT"
)
obj.cd8.int$cellstate_manual <- factor(
  plyr::mapvalues(
    x = as.character(Idents(obj.cd8.int)),
    from = names(cluster_labels),
    to = unname(cluster_labels)
  ),
  levels = levels_order
)
celltype_cols <- c(
  CD8_Naive_TCM_like = "#E6C229",
  CD8_EarlyMemory_like = "#FFF3A3",
  CD8_StemLike_Memory = "#9BE3E0",
  CD8_Transitional_EffectorMemory = "#F299C1",
  CD8_Activated = "#E39A2D",
  CD8_Activated_PD1pos = "#2D8C7F",
  CD8_ChronicallyActivated_CXCL13pos = "#7A1C3B",
  CD8_TEMRA = "#6A3DFF",
  CD8_NK_Like = "#4B1FA6",
  MAIT = "#39D353"
)
cols_pools <- c(
  "High PDCD1 controls" = "#3F5A8A",
  "Highly expanded controls" = "#8E5EA2",
  "Low PDCD1 controls" = "#6BAED6",
  "NeoAg_pools" = "#E45756",
  "NA" = "#7F7F7F"
)
```

# umaps

```{r}
# --- Seurat clusters ---
plot.a.cd8 <- DimPlot(obj.cd8.int,
        reduction = "umap",
        group.by = "seurat_clusters",
        label = T,
        repel = F,
        seed = 123,
        label.size = 5) + 
  theme(aspect.ratio = 1)
# --- Phenotype or cellular state ---
plot.b.cd8 <- DimPlot(obj.cd8.int,
        reduction = "umap",
        group.by = "cellstate_manual",
        cols = celltype_cols,
        label.size = 5) + 
  theme(aspect.ratio = 1)
# --- Validated TCR IDs ---
plot.c.cd8 <- DimPlot(obj.cd8.int,
        reduction = "umap",
        group.by = "TCR.ID",
        cols = tcr_colors,
        pt.size = ifelse(is.na(obj.cd8.int$TCR.ID), 0.25, 1),
        alpha = ifelse(is.na(obj.cd8.int$TCR.ID), 0.25, 1),
        label.size = 5) + 
  theme(aspect.ratio = 1)
# --- Validated TCR IDs (pools) ---
plot.d.cd8 <- DimPlot(obj.cd8.int,
        reduction = "umap",
        group.by = "Pools",
        cols = cols_pools,
        pt.size = ifelse(is.na(obj.cd8.int$TCR.ID), 0.25, 1),
        alpha = ifelse(is.na(obj.cd8.int$TCR.ID), 0.25, 1),
        label.size = 5
        ) + 
  theme(aspect.ratio = 1)
plots_list[["Umap_CD8_TCRs"]] <- wrap_plots(plot.a.cd8, plot.b.cd8, plot.c.cd8, plot.d.cd8, ncol = 2)
print(plots_list[["Umap_CD8_TCRs"]])
```

TCRs identified from TIL expansion pools (5_TIL-15_IP-TIL_ and 7_TIL-15_IP-TIL_).

```{r}
til_pattern <- "5_TIL-15_IP-TIL_|7_TIL-15_IP-TIL_"
colors <- c(
  "TIL_pool_reactive"   = "red3",
  "TIL_pool_non_tested" = "black",
  "No"                  = "gray70"
)
# --- Top-100% ---
res_top100 <- annotate_plot_til_pools(
  obj = obj.cd8.int,
  pool_col = "pool",
  out_col = "til_pools",
  plot_name = "Umap_CD8_til_pools_top100"
)
tcr_states_top100 <- res_top100$tcr_states
# --- Top-70% ---
res_top70 <- annotate_plot_til_pools(
  obj = obj.cd8.int,
  pool_col = "pool_top70",
  out_col = "til_pools_top70",
  plot_name = "Umap_CD8_til_pools_top70"
)
tcr_states_top70 <- res_top70$tcr_states
```

# stats

Calculate the percentage of reactive cells and TCRs in the overall CD8+ T cell population and the CD8+CD39+ T cell population.

```{r, warning=FALSE}
valid_tcr <- function(x) {
  x[!is.na(x) & x != "no_tcr" & x != ""]
}
# --- Cell level ---
# CD8+
total_cells <- dim(obj.cd8.int)[2]
reactive_cells <- sum(obj.cd8.int$Reactive == "Yes", na.rm = TRUE)
pct_overall <- (reactive_cells / total_cells) * 100
# CD8+CD39+
obj.cd8.int.sub <- subset(obj.cd8.int, subset = tag.compartment == "CD8+CD39+")
total_cells <- dim(obj.cd8.int.sub)[2]
reactive_cells <- sum(obj.cd8.int.sub$Reactive == "Yes", na.rm = TRUE)
pct_cd8cd39 <- (reactive_cells / total_cells) * 100
# --- TCR level ---
# CD8+
total_tcr <- length(unique(valid_tcr(obj.cd8.int$cdr3_vj)))
reactive_tcr <- obj.cd8.int@meta.data %>%
  dplyr::filter(Reactive == "Yes") %>%
  dplyr::pull(cdr3_vj) %>%
  valid_tcr() %>%
  unique() %>%
  length()
pct_overall_tcr <- (reactive_tcr / total_tcr) * 100
# CD8+CD39+
total_tcr <- length(unique(valid_tcr(obj.cd8.int.sub$cdr3_vj)))
reactive_tcr <- obj.cd8.int.sub@meta.data %>%
  dplyr::filter(Reactive == "Yes") %>%
  dplyr::pull(cdr3_vj) %>%
  valid_tcr() %>%
  unique() %>%
  length()
pct_cd8cd39_tcr <- (reactive_tcr / total_tcr) * 100
# --- Summary table ---
summary_table <- data.frame(
  Subset = c("CD8+", "CD8+CD39+"),
  Percentage_Cells_reactive = c(pct_overall, pct_cd8cd39),
  Percentage_TCRs_reactive  = c(pct_overall_tcr, pct_cd8cd39_tcr)
)
print(summary_table)
```

# save

```{r}
# --- Save plots ---
dir.create(file.path(datapath, sample, folder_data, "plots"), recursive = FALSE)
for (i in names(plots_list)) {
  outfile <- file.path(datapath, sample, folder_data, "plots", paste0(i, ".pdf"))
  tryCatch({
    pdf(file = outfile, width = 297 / 15, height = 210 / 15)
    print(plots_list[[i]])
    dev.off()
  }, error = function(e) {
    message("Error en plot: ", i)
    message("  ", e$message)
    if (dev.cur() != 1) {
      try(dev.off(), silent = TRUE)
    }
  })
}
# --- Save frequency tables ---
dir.create(file.path(datapath, sample, folder_data, "frequencies"), recursive = F)
write.table(tcr_states_top70, file.path(datapath, sample, folder_data, "frequencies", "CD8_til_pools_top70.tsv"), sep = "\t", dec = ",", row.names = FALSE, quote = FALSE)
write.table(tcr_states_top100, file.path(datapath, sample, folder_data, "frequencies", "CD8_til_pools_top100.tsv"), sep = "\t", dec = ",", row.names = FALSE, quote = FALSE)
# --- Save Seurat object ---
saveRDS(obj, file.path(datapath, sample, folder_data, "seurat/obj_filt_final.rds"))
saveRDS(obj.cd8.int, file.path(datapath, sample, folder_data, "seurat/obj_filt_cd8_final.rds"))
```